Bayesian exponential random graph modeling of KWL Burial network

### 2020-10-26 analysis/paper/paper.Rmd
delete code chunks in the paper rmd file

```{r tidy-data}
# combine some periods
burial_three_period_age_tidy <-
  burial %>%
  rename(burial_label = ID) %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>%
  mutate(Gold_leaf = ifelse(Gold_leaf == "shatter", "1", Gold_leaf),
         Stoneware = ifelse(Stoneware == "base", "1", Stoneware),
         Stamped_ceramic = ifelse(Stamped_ceramic == "cluster", "1", Stamped_ceramic)) %>%
  mutate_at(21:ncol(.), as.numeric) %>%
  janitor::remove_empty(which = "cols") %>%
  mutate(total = rowSums(.[c(21:48, 50, 55, 56)], na.rm = TRUE)) %>% #prestige goods
  mutate(Porcelain = rowSums(.[c(40:48, 55, 56)], na.rm = TRUE)) %>% #B&W, porcelain, Anping, kendi
  mutate(Porcelain = ifelse(Porcelain == 0, NA, Porcelain)) %>%
  mutate(quantity = case_when(
    total == 0 ~ "none",
    total > 0 & total <= 7 ~ "low",
    total > 7 & total < 100 ~ "medium",
    total >= 100 ~ "high",
    TRUE ~ "")) %>% # the classification is based on the result of histogram
  mutate(Age_scale = case_when(
    `Age` %in% c("1","2") ~ "0-12",
    `Age` == "3" ~ "12~20",
    `Age` %in% c("4","5","6","7","8") ~ "+20",
    TRUE ~ "NA")) %>%
  mutate(gender = case_when(
    `Gender` %in% c("1","2") ~ "male",
    `Gender` %in% c("3","4") ~ "female",
    TRUE ~ "NA")) %>%
  mutate(ritual = case_when(
    `Stamped_ceramic` == "2" ~ "two pots",
    `Stamped_ceramic` == "1" ~ "one pot",
    TRUE ~ "NA")) %>%
  mutate(Gold_bead_low = ifelse(Golden_bead == 1, 1, NA),
         Gold_bead_med = ifelse(Golden_bead > 1 & Golden_bead <10, 1, NA),
         Gold_bead_high = ifelse(Golden_bead > 10, 1, NA),
         Agate_bead_low = ifelse(Agate_bead == 1, 1, NA),
         Agate_bead_med = ifelse(Agate_bead > 1 & Agate_bead <10, 1, NA),
         Agate_bead_high = ifelse(Agate_bead > 10, 1, NA),
         `Indo-Pacific_bead_low` = ifelse(`Indo-Pacific_bead` < 100, 1, NA),
         `Indo-Pacific_bead_med` = ifelse(`Indo-Pacific_bead` > 100 & `Indo-Pacific_bead` < 900, 1, NA),
         `Indo-Pacific_bead_high` = ifelse(`Indo-Pacific_bead` > 900, 1, NA)) %>% #based on the result of histogram
  select(burial_label,
         Phase,
         Age_scale,
         gender,
         ritual, # consider to remove if not very informative
         Gold_bead_low,
         Gold_bead_med,
         Gold_bead_high,
         Agate_bead_low,
         Agate_bead_med,
         Agate_bead_high,
         Porcelain, #prestige good
         Gold_leaf, #prestige good
         fish_shape_knit, #prestige good
         quantity)

# number of each phase
burial_three_period_age_number <-
  burial_three_period_age_tidy %>%
  count(Phase)

pre_E <- burial_three_period_age_number$Phase[4]
post_E <- burial_three_period_age_number$Phase[3]
chi <- burial_three_period_age_number$Phase[1]
```

```{r prep-network-object-pre}
# filter pre burials
burial_pre <-
  burial_three_period_age_tidy %>%
  filter(Phase == "pre") %>%
  janitor::remove_empty(which = "cols")

# create node list using burial index by phase
#--------------------pre--------------------------------
nodes_pre <-
  burial_three_period_age_tidy %>%
  filter(Phase == "pre") %>%
  select(burial_label) %>%
  rowid_to_column("id")

# pair wise combinations for burials as index for later map function
library(gtools)
burial_comb_pre = combinations(length(nodes_pre$burial_label), #t(combn(nodes_pre$burial_label, 2))
                               2, nodes_pre$burial_label,
                               repeats = TRUE)
colnames(burial_comb_pre) = c("burial_1", "burial_2")
burial_comb_pre = as_tibble(burial_comb_pre)

# create list for each burial that contains the burial good types and their counts
edge_list_pre <-
  burial_three_period_age_tidy %>%
  select(burial_label, 6:14) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  #mutate(burial_connection = rep(unique(burial_label), length.out = length(burial_label)))
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_pre <-
  map2(burial_comb_pre$burial_1,
       burial_comb_pre$burial_2,
       ~bind_cols(
         edge_list_pre %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_pre %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count...3, count...6)))

# count of ornament types in common between each pair of burials
common_counts_vct_pre <- map_int(common_counts_lst_pre, ~sum(!is.na(.x$common_counts)))

burial_comb_with_common_counts_pre <-
  burial_comb_pre %>%
  mutate(common_counts = common_counts_vct_pre) %>%
  mutate(common_counts = ifelse(burial_1 == burial_2, 0, common_counts)) # use 0 for self loop

# change label to ids for node linking
edges_pre <-
  burial_comb_with_common_counts_pre %>%
  left_join(nodes_pre, by = c("burial_1" = "burial_label")) %>%
  rename(from = id) %>%
  left_join(nodes_pre, by = c("burial_2" = "burial_label")) %>%
  rename(to = id)

edges_for_network_pre <-
  select(edges_pre, from, to, common_counts) %>%
  filter (!common_counts == 0) # remove rows with no goods in common
# %>% mutate(common_counts = ifelse(common_counts > 1, 1, common_counts)) # for unweighted network
```

```{r prep-network-object-post}
# filter post burials
burial_post <-
  burial_three_period_age_tidy %>%
  filter(Phase == "post") %>%
  janitor::remove_empty(which = "cols") %>%
  filter(quantity != "none") # remove burial without burial goods

# create node list
nodes_post <-
  burial_three_period_age_tidy %>%
  filter(Phase == "post") %>%
  filter(quantity != "none") %>%
  select(burial_label) %>%
  rowid_to_column("id")

# pair wise combinations for burials as index for later map function
library(gtools)
burial_comb_post = combinations(length(nodes_post$burial_label),
                               2, nodes_post$burial_label,
                               repeats = TRUE)
colnames(burial_comb_post) = c("burial_1", "burial_2")
burial_comb_post = as_tibble(burial_comb_post)

# create list for each burial that contains the burial good types and their counts
edge_list_post <-
  burial_three_period_age_tidy %>%
  select(burial_label, 6:14) %>% # need to change for each exploration
  pivot_longer(-burial_label, names_to = "goods", values_to = "count") %>%
  group_by(burial_label) %>%
  nest()

# pair-wise list
common_counts_lst_post <-
  map2(burial_comb_post$burial_1,
       burial_comb_post$burial_2,
       ~bind_cols(
         edge_list_post %>%
           filter(burial_label == .x) %>%
           unnest(data),
         edge_list_post %>%
           filter(burial_label == .y) %>%
           unnest(data)) %>%
         rowwise() %>%
         mutate(common_counts = sum(count...3, count...6)))

# count of ornament types in common between each pair of burials
common_counts_vct_post <- map_int(common_counts_lst_post, ~sum(!is.na(.x$common_counts)))

burial_comb_with_common_counts_post <-
  burial_comb_post %>%
  mutate(common_counts = common_counts_vct_post) %>%
  mutate(common_counts = ifelse(burial_1 == burial_2, 0, common_counts)) # use 0 for self loop

# change label to ids for node linking
edges_post <-
  burial_comb_with_common_counts_post %>%
  left_join(nodes_post, by = c("burial_1" = "burial_label")) %>%
  rename(from = id) %>%
  left_join(nodes_post, by = c("burial_2" = "burial_label")) %>%
  rename(to = id)

edges_for_network_post <-
  select(edges_post, from, to, common_counts) %>%
  filter (!common_counts == 0) # remove rows with no goods in common
```

@@

```{r network-ggraph}
# network analysis using ggraph pkg
library(tidygraph)
library(ggraph)
# pre-Euro
relation_tidy_pre <- tbl_graph(nodes = nodes_pre,
                               edges = edges_for_network_pre,
                               directed = FALSE)

relation_tidy_pre %>%
  activate(edges) %>%
  arrange(desc(common_counts))

# plot
ggraph(relation_tidy_pre, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()

# post-Euro
relation_tidy_post <- tbl_graph(nodes = nodes_post,
                               edges = edges_for_network_post,
                               directed = FALSE)

relation_tidy_post %>%
  activate(edges) %>%
  arrange(desc(common_counts))

# plot
ggraph(relation_tidy_post, layout = "graphopt") +
  geom_node_point() +
  geom_edge_link(aes(width = common_counts), alpha = 0.8) +
  scale_edge_width(range = c(0.2, 2)) +
  geom_node_text(aes(label = burial_label), repel = TRUE) +
  labs(edge_width = "common item") +
  theme_graph()
```

```{r network-object}
library(network)

burial_network_pre <-
  network(edges_for_network_pre, # the network object
          vertex.attr = nodes_pre, # node list
          directed = FALSE, # specify whether the network is directed
          ignore.eval = FALSE, # FALSE = weighted
          loops = FALSE, # do we allow self ties (should not allow them)
          matrix.type = "edgelist") # the type of input

burial_network_post <-
  network(edges_for_network_post,
          vertex.attr = nodes_post,
          directed = FALSE,
          ignore.eval = FALSE, 
          loops = FALSE, 
          matrix.type = "edgelist") 
```

@@

```{r add-node-attribute-color}
# prepare notwork objects for Bayesian ERGMs 
library(statnet)
library(Bergm)

# Add attributes to the network object burial_network_pre
set.vertex.attribute(burial_network_pre, "quantity", burial_pre$quantity)
set.vertex.attribute(burial_network_pre, "age", burial_pre$Age_scale)
set.vertex.attribute(burial_network_pre, "gender", burial_pre$gender)

# plot
set.seed(30)
quantity <- get.vertex.attribute(burial_network_pre, "quantity")
age <- get.vertex.attribute(burial_network_pre, "age")

plot(burial_network_pre,
     displaylabels = TRUE,
     vertex.col = "quantity",
     vertex.cex = degree(burial_network_pre, cmode = 'indegree')/5, #size nodes
     pad = 1)

legend("topleft",
       col = c(2, 3, 1, 4), # need to adjust each time
       pch    = 20,
       legend = unique(quantity), # age
       title  = 'Burial good counts')

#? can't get the items in legend in order, tried as.factor and set their level but does not work
#? can't match color with categories, using an odd method at the moment

# Add attributes to the network object burial_network_post
set.vertex.attribute(burial_network_post, "quantity", burial_post$quantity)
set.vertex.attribute(burial_network_post, "age", burial_post$Age_scale)
set.vertex.attribute(burial_network_post, "gender", burial_post$gender)

# plot
set.seed(30)
quantity <- get.vertex.attribute(burial_network_post, "quantity")
age <- get.vertex.attribute(burial_network_post, "age")
ID <- get.vertex.attribute(burial_network_post, "burial_label") # not sure how to get id on the network plot

plot(burial_network_post,
     displaylabels = TRUE,
     vertex.col = "quantity",
     vertex.cex = degree(burial_network_post, cmode = 'indegree') / 13, #size nodes to their in-degree
     #vertex.sides = ifelse(burial_network_pre %v% "", 4, 50),
     pad = 1)

legend("topleft",
       col = c(4, 3, 2, 1), # need to adjust each time
       pch    = 20,
       legend = unique(quantity),
       title  = 'Burial good counts')
```

@@

```{r network-ERGMs}
# model 1
model.ergm <- burial_network_pre ~
  edges + 
  density +
  triangle 
summary(model.ergm)

# model 2 considers cluster and degree
model.2 <- burial_network_pre ~ 
  edges + # density
  gwesp(0.2, fixed = TRUE) +  # transitivity(cohesion; triangle)
  gwdegree(0.8, fixed = TRUE)  # popularity(degree; star), the frequency distribution for nodal degrees
summary(model.2)

# model 3 with node attributes
model.3 <- burial_network_pre ~ edges +  # the overall density of the network
  nodematch('quantity') + # quantity-based homophily
  nodematch('age') +
  nodematch('gender') +
  gwesp(0.75, fixed = TRUE) +    # transitivity
  gwnsp(0.75, fixed = TRUE) +
  gwdegree(0.2, fixed = TRUE) # popularity
summary(model.3)
```

@@

```{r Bayesian-ERGMs}
# Specify a prior distribution: normal distribution (low density and high transitivity)
# need to observe the graph and select suitable prior 
prior.mean <- c(2, 0, 0, 0, 5, 0, 0) 
# follow Alberto Caimo et al. (2015) hospital example
prior.sigma <- diag(3, 7) # covariance matrix structure
# where the dimension d corresponds to the number of parameters, 𝜇 is mean vector and Σprior is a d × d covariance matrix.

# Estimated posterior means, medians and 95% credible intervals for Models.3
# bergmM: Bayesian exponential random graphs models under missing data using the approximate exchange algorithm. Use this because of missing data for age variable
parpost <- bergmM(model.3,
                  prior.mean  = prior.mean,
                  prior.sigma = prior.sigma,
                  burn.in     = 200, # burn-in iterations for every chain of the population, drops the first 200
                  main.iters  = 2000, # iterations for every chain of the population
                  aux.iters   = 10000, # MCMC steps used for network simulation
                  nchains     = 8, # number of chains of the population MCMC
                  gamma       = 0.2) # scalar; parallel adaptive direction sampling move factor, acceptance rate

# summary of parameter estimation
summary(parpost)

# plot the parameter posterior distribution
plot(parpost)
```

@@

```{r Bayesian-gof}
# Model assessment, Bayesian goodness of fit diagnostics:
bgof(parpost,
     aux.iters = 10000,
     n.deg     = 15,
     n.dist    = 15,
     n.esp     = 15)

# An estimated ERGM is fitting perfectly a certain observed network if the red line falls inside this interval
# a result that is very difficult to obtain in practice (Caimo 2017)
```

>>>>

```{r tidy-data}
# number of each phase
burial_phases_sample_size <-
  burial %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>% 
  count(Phase)

pre_E <- burial_phases_sample_size %>% filter(Phase == "pre") %>% pull()
post_E <- burial_phases_sample_size %>% filter(Phase == "post") %>% pull()
chi <- burial_phases_sample_size %>% filter(Phase == "chi") %>% pull()
```

@@

@@

@@

```{r network-ERGMs}
```

@@

```{r Bayesian-ERGMs}
```

@@

```{r Bayesian-gof}
```
